---
title: "R programming basics (part 2a)"
author: "Prof. Francesca Finotello"
editor: visual
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    toc-depth: 3
    toc-floating: true   # TOC stays visible while scrolling
---

# Introduction

In this exercise set, you will learn the basics of programming in R by creating functions, using loops, and working with conditional execution.

# Exercises

## Exercise 1

Write a function `square(x)` that takes as input a number, specified by the `x` parameter, and computes x to the power of 2 (i.e., $x^2$), and returns this result using the `return` function.

<details>

<summary>Click to see the solution code</summary>

**Solution:**

```{r}
square <- function(x) {
    x <- x^2
    return(x)
}
```

</details>

Compute `5^2` with your function and verify it returns `25`.

## Exercise 2

Write a second function `powfun(x, y)` that computes "x to the power of y" (i.e., $x^y$) from the input data `x` and `y`.

Verify that the results of `powfun(x = 6, y = 2)` correspond to the ones obtained with `square(x = 6)`.

## Exercise 3

Write a function called `fahrenheit_to_celsius` that converts temperatures from degrees Fahrenheit to degrees Celsius according to the following formula:

$\text{Celsius} = (\text{Fahrenheit} - 32) \times 5 / 9$

Check if 30 degrees Fahrenheit correspond to a temperature in degrees Celsius that is above 0.

<details>

<summary>Click to see the solution code</summary>

**Solution:**

```{r}
fahrenheit_to_celsius <- function(temp_f) {
  temp_c <- (temp_f - 32) * 5 / 9
  return(temp_c)
}
```

</details>

## Exercise 4

Write a second function called instead `celsius_to_fahrenheit` that converts temperatures from degrees Celsius to degrees Fahrenheit according to the following formula:

$\text{Fahrenheit} = \text{Celsius} \times 9 / 5 + 32$

## Exercise 5

Write a function called `water_state` that takes as input the temperature `temp` in degrees Celsius and returns a *character* indicating if the water is in a **solid** (i.e., frozen, if `temp<0`), **gas** (if `temp>=100`), or **liquid** state. The final label should be returned with the `return` function. *Hint:* use `if`, `else if`, and `else` conditional statements.

```{r, include=FALSE, echo=FALSE}
water_state <- function(temp) {
    if (temp < 0) {
        res <- "solid"
    } else if (temp < 100) {
        res <- "liquid"
    } else {
        res <- "gas"
    }
    return(res)
}
```

Check that the function returns the following results:

```{r}
water_state(-1)
water_state(0)
water_state(10)
water_state(100)
water_state(120)
```

## Exercise 6

The definition of the mean of a vector of numbers `x` is as follows:

$$
\bar{x} = \frac{\sum_{i=1}^{n} x_i}{n} 
$$ 

This corresponds to summing up all numbers in `x` and dividing this sum by the number of elements in `x`.

Imagine we have the following vector

```{r}
x <- c(2, 4, 10, 0, 2, 4)
```

We can compute the mean directly with the `mean` function, or as the ratio between the sum and the number of its elements:

```{r}
mean(x)
sum(x) / length(x)
```

Let's imagine instead that we want to define a new function called `my_mean` which walks through the vector one element at a time and, at each iteration updates the current sum and count of the vector elements.

For this, within the function we would need two elements to keep track of the number of elements in (`x_count`) and the sum of the element values (`x_sum`). These two elements should be initialized at the beginning of the function (**within** the function), setting them to 0 (i.e., no vector element has been considered yet):

```{r}
x_count <- 0
x_sum <- 0
```

We can then use a `for` loop to walk through all the positions of the input vector `x`, from 1 to `length(x)`. At every iteration, we should update `x_count` to reflect that we count an additional (`+1`) element, and `x_sum` to add the value of the current element of `x`. At the end of the loop, we can compute the ratio between `x_sum` and `x_count`, and return the results with the `return` function. Define such the `my_mean` function.

<details>

<summary>Click to see the solution code</summary>

**Solution:**

```{r, include=TRUE}
my_mean <- function(x) {
    x_sum <- 0
    x_count <- 0
    for (i in 1:length(x)) {
        x_sum <- x_sum + x[i]
        x_count <- x_count + 1
    }
    return(x_sum / x_count)
}  
```

</details>

Verify that it gives the same results of the official `mean` function, as tested below:

```{r}
x <- seq(-21, 100)
y <- c(2, 3, 7, 2, 0, 10)
my_mean(x); mean(x)
my_mean(y); mean(y)
```

## Exercise 7

Let's compute the `my_min` function, that takes as input a vector `x` of numbers, uses a loop to iterate across the elements of the vector to identify its smallest value.

The function should use a variable called `x_min` to be used to store the current minimum. We need to initialize using the biggest value possible, otherwise we risk that no element in the vector is chosen:

```{r}
x_min <- Inf
```

Then we should use a loop to walk through all the elements of the vector. For each iteration, we check if the current element is smaller than `x_min`: if yes, we save the current value in `x_min`; otherwise, we disregard the current element and move to the next one. At the end of the loop, the function should `return` the minimum saved in `x_min`.

Implement the `my_min` function as suggested above (`min` and `which.min` cannot be used!) and verify that it gives the same results of the official `min` function, as tested below:

```{r, include=FALSE}
my_min <- function(x) {
    x_min <- Inf
    for (i in 1:length(x)) {
        if (x[i] < x_min) x_min <- x[i]
    }
    return(x_min)
}  
```

```{r}
x <- seq(-21, 100)
y <- c(2, 3, 7, 2, 0, 10)
my_min(x); min(x)
my_min(y); min(y)
```

## Exercise 8

Without executing the code, what value does `i` have after running the code below?

1.  100
2.  4
3.  2
4.  None of the above

```{r, echo=TRUE}
y <- c(-1, 100, 2, 4)

for (i in 1:length(y)) {
  if (y[i] == 100) {
    break
  }
}
```

<details>

<summary>Click to see the answer</summary>

**Answer:** 3. 2

</details>

## Exercise 9

Without executing the code, what value does `i` have after running the code below?

1.  9
2.  6
3.  4
4.  None of the above

```{r, echo=TRUE}
j <- 0
x <- c(0, -1, 3, -4, 4, 2)

for (i in 1:length(x)) {
  if (x[i] > 0) {
    j <- j + x[i]
  }
}
```

<details>

<summary>Click to see the answer</summary>

**Answer:** 2. 6

</details>

## Exercise 10

Without executing the code, what value does `warn` have after running the code below?

1.  Hypoglycemia
2.  Hyperglycemia
3.  Normoglycemia
4.  NA

```{r, echo=TRUE}

glycemia <- 120
hypo_thresh <- 70; hyper_thresh <- 130 # mg/dL
warn <- NA
if (glycemia < hypo_thresh) {
  warn <- "Hypoglycemia"
}  else  if (glycemia > hyper_thresh) {
  warn <- "Hyperglycemia"
} else {
  warn <- "Normoglycemia"
}

```

<details>

<summary>Click to see the answer</summary>

**Answer:** 3. "Normoglycemia"

</details>
